Software architectures to integrate workflow engines in science gateways

### 2016-10-12 answers.tex
Even more answers to the reviewers

\todo{think of this}

@@

\todo{just remove CSGF, Airvata and pgrade from the table, keeping the
  exponents. But we don't have examples of gateways using Airvata,
  still need to be found.}
\end{answer}

@@

\begin{answer}
As suggested by the reviewer, the presentation of these Figures has been gathered in a separate section, the new section 3.
\end{answer}

@@

\begin{answer}
  We agree with the reviewer that functionality is important. This was
  exactly the goal of the ``Specific features'' evaluation
  criterion. We renamed this criterion ``Functionality'' as we find it a much better name indeed. We also renamed O$_1$ to F$_1$ and O$_2$ to F$_2$.
\end{answer}

@@

\todo{complexity is used as a measure of reliability}

\begin{review}
I therefore think that the metrics defined only provide a very high level structure for an evaluation of the different software architectures and they should discuss other factors and scenarios that can affect development. Some of these limitations are discussed in section 4.2 but I think an extra section discuss these before the definition of the metrics would be very helpful to set the scene correctly for the evaluation. Otherwise, the reader is left asking a number of questions during the evaluation section. I also think that a few more metrics e.g. "functionality" could be added to provide a more balanced view of the different approaches. Another useful extension of the paper could be to normalize the results against the functionality each approach provides (perhaps in a separate table) - this may help to provide a more balanced table that factors in functionality.
\end{review}

>>>>

\begin{answer}
  We agree with the reviewer that using WSRP (Web Services for Remote
  Portlets) changes the architecture. When portlets are deployed on
  different hosts, we think that the architecture becomes service
  invocation as the portlets then behave as individual software
  components according to our definitions. We added the following comment as a footnote in the presentation of tight integration in section 2.2:
\revised{Note that the
    particular case of portlets can be a bit ambiguous: when portlets
    are deployed remotely, using a protocol such as Web Services for
    Remote Portlets (WSRP), they become independent software
    components according to our definition because they could be
    operated by different teams on different hardware systems. In this
    case, the system should be classified in the service invocation
    architecture since the portlets actually become Web
    Services. The same comment holds for remote object
    invocation.}
\end{answer}

@@

We totally agree with the reviewer that Apache Airavata and
WS-PGrade/gUSE are gateway frameworks. This is how they are presented
in the text. Instead of providing an incomplete list of gateways that
use these frameworks in Table 1, we now refer explicitly to ``gateways
based on Apache Airvata'', ``gateways based on WS-PGRADE/gUSE'' and
``gateways based on the Catania Science Gateway
Framework''. Consistently, the table doesn't mention the individual
gateways based on these frameworks any more.
\end{answer}

@@

\begin{answer}
  As suggested by the reviewer, the presentation of these Figures has
  been gathered in a separate section which is now the new section 3.
\end{answer}

@@

\begin{answer}
  We agree with the reviewer that functionality is important. This was
  exactly the goal of the ``Specific features'' evaluation
  criterion. We renamed this criterion ``Functionality'' as we find it
  a much better name indeed. We also renamed O$_1$ to F$_1$ and O$_2$
  to F$_2$.

  However, we don't think that ``the ability to pass bundles'' should
  be listed as a metric in the functionality criterion, since this
  feature remains quite specific to the particular context mentioned
  by the reviewer. 
\end{answer}

@@

\begin{answer}
  We totally agree with the reviewer. However, we think that this
  issue cannot be addressed at the pure architectural level as it
  requires accurate technical information about systems. In some
  systems, the science gateway may become a massive piece of
  unmaintainable code as described by the reviewer, and in some other
  cases, it may have a neat internal architecture that allows for easy
  integration of workflow engines. Once again, we believe that the
  answer to this issue is about assigning weights to the different
  components and interactions, which is now possible through the
  public spreadsheet. For instance, if the science gateway code is
  known to be ``untouchable'' due to a cumbersome, then a high weight
  should be put on the science gateway component. 
\end{answer}

\begin{review}
  I therefore think that the metrics defined only provide a very high
  level structure for an evaluation of the different software
  architectures and they should discuss other factors and scenarios
  that can affect development. 
\end{review}

\begin{answer}
  The metrics provide an abstract framework for evaluating
  architectures independently from any particular system. The goal is
  to abstract architectural discussions from the particular technical
  context of a real system. Some degree of system-specificity can now
  be introduced by assigning weights to interactions and components
  through our public spreadsheet, which we think provides a way to apply our
  analysis to study real systems more precisely.
\end{answer}

\begin{review}
Some of these limitations are discussed in section 4.2 but I think an
extra section discuss these before the definition of the metrics would
be very helpful to set the scene correctly for the
evaluation.  Otherwise, the reader is left asking a number of questions
during the evaluation section. 
\end{review}

\begin{answer}
We added the following paragraphs to the introduction to section 3:

\revised{Note that the metrics defined hereafter aim at providing a framework
to evaluate architectures independently from any particular system,
regardless of any technical constraint such as the use of a specific
programming language or the internal architecture of the science
gateway. The goal is to abstract architectural discussions from the
particular technical context of a real system. Some degree of
system-specificity can be introduced by assigning weights to
interactions and components through a public spreadsheet available at
\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
and interactions" contains the weight given to these entities in the
metrics and can be used to produce new versions of
Table 2 that fit the particular technical context
of a real system. For instance, a high weight could be assigned to the
science gateway if its internal architecture is such that it can
hardly be extended in practice.}

\revised{In addition, all the metrics in a criterion contribute
  equally to the global score used for the criterion. Column B of
  sheet ``Metrics'' of the public spreadsheet may be used to assign
  different weights to each metric to adjust the specific context of a
  study.}
\end{answer}

\begin{review}
I also think that a few more metrics
e.g. "functionality" could be added to provide a more balanced view of
the different approaches.

 Another useful extension of the paper could
be to normalize the results against the functionality each approach
provides (perhaps in a separate table) - this may help to provide a
more balanced table that factors in functionality.
\end{review}

\begin{answer}
  As explained previously, the last evaluation criterion, now named
  ``Functionality'' aims at this purpose. We hope it clarifies that
  functionality is actually part of the analysis. 
\end{answer}

### 2016-10-12 paper.tex
Even more answers to the reviewers

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation (``Service+Service'').]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow conversion. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{ardizzone2012decide},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, Neuroscience Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$, GridSFEA~\cite{Elts2010}
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Conversion & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where applications, users and other
resources are stored.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine,
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

@@

\begin{figure*}
\centering
\def\svgwidth{1.5\columnwidth}
\input{figures/VIP.pdf_tex}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
  performed by the tasks and embedded in their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
\correction{functionality}. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each criterion, a global score is
computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score
for a criterion are comparable, so that they can be combined
meaningfully. \correction{All the metrics in a criterion contribute
  equally to the global score. Should the metrics be weighted
  differently in a particular system, the reader is invited to produce
  his/her own version of the Table using the public spreadsheet
  available at \url{https://frama.link/LrPRsCQp} (see column B of
  sheet ``Metrics'').}

>>>>

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation (``Service+Service'').]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow conversion. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          \correction{Gateways based on the} Catania Science Gateway Framework~\cite{ardizzone2012decide},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          LONI Pipeline Environment~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
             \correction{Gateways based on} Apache Airvata~\cite{marru2011apache}, HubZero with Pegasus~\cite{CPE:CPE3257}, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13},           \correction{Gateways based on the} WS-PGRADE/gUSE framework~\cite{Kacsuk2012}, GridSFEA~\cite{Elts2010}
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}, CBRAIN and FSL\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling}), HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Conversion & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines.}
\label{table:system-classification}
\end{table*}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model in a
Ruby on Rails application\footnote{\correction{Note that the
    particular case of portlets can be a bit ambiguous: when portlets
    are deployed remotely, using a protocol such as Web Services for
    Remote Portlets (WSRP), they become independent software
    components according to our definition because they could be
    operated by different teams on different hardware systems. In this
    case, the system should be classified in the service invocation
    architecture since the portlets actually become Web
    Services. The same comment holds for remote object
    invocation.}}. The workflow engine and the science gateway usually
share a database where applications, users and other resources are
stored.  In this model, task and data control are both initiated from
the science gateway. Interactions \texttt{b} and \texttt{c$_2$} are
initiated from the workflow engine, while \texttt{c$_1$} comes from
other parts of the science gateway, for instance a data management
interface. As in any other model, the installation of new workflows in
the science gateway (\texttt{a$_2$}) and infrastructure
(\texttt{a$_1$}) is done by an administrator, for security
reasons. This is the model adopted in the Catania Science Gateway
Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

\correction{This section illustrates the architectural patterns on a
  few real systems. The architecture diagrams presented are the
  original ones as published by the system developers, which we
  annotated using the interactions and components in
  Figure~\ref{fig:architectures}. }

@@

\begin{figure*}
\centering
%\def\svgwidth{1.5\columnwidth}
\includegraphics[width=0.9\textwidth]{figures/VIP}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds to
  interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP, workflow data transfers are
  performed by the tasks and embedded in their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
\correction{functionality}. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each criterion, a global score is
computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score
for a criterion are comparable, so that they can be combined
meaningfully.

\correction{Note that the metrics defined hereafter aim at providing a framework
to evaluate architectures independently from any particular system,
regardless of any technical constraint such as the use of a specific
programming language or the internal architecture of the science
gateway. The goal is to abstract architectural discussions from the
particular technical context of a real system. Some degree of
system-specificity can be introduced by assigning weights to
interactions and components through a public spreadsheet available at
\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
and interactions" contains the weight given to these entities in the
metrics and can be used to produce new versions of
Table~\ref{table:evaluation} that fit the particular technical context
of a real system. For instance, a high weight could be assigned to the
science gateway if its internal architecture is such that it can
hardly be extended in practice.}

\correction{In addition, all the metrics in a criterion contribute
  equally to the global score used for the criterion. Column B of
  sheet ``Metrics'' of the public spreadsheet may be used to assign
  different weights to each metric to adjust the specific context of a
  study.}
